Synthesising Recursive Functions for First-Order Model Counting

### 2023-03-07 doc/paper/supplementary_material.tex
Describe one more instance
Clean up supplementary material

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usepackage{siunitx}
\usepackage{forest}
\usepackage[inline]{enumitem}
\usepackage{xr}
\externaldocument{paper}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}[theorem]
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

\title{Synthesising Recursive Functions for First-Order Model Counting}

% TODO (at the very end)
% put the appendix in a separate file
% grammarly the whole thing
% read the formatting instructions and their LaTeX source (email font?)
% make sure there are no overflow errors
% clean up unnecessary things from the header (e.g., packages and definitions)
%  make sure the result table (as well as other floating objects) are on the right pages

\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore, Singapore\\
$^2$University of Edinburgh, Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}

@@

@@

\begin{theorem}[Correctness of $\Reff$]
  Let $\phi$ be the formula used as input to \cref{alg:trycache}. Let $\psi$ be
  any formula selected on \cref{line:selectformula} of the algorithm s.t.\
  $\rho \ne {\normalfont\texttt{null}}$ on \cref{line:rho}. Let $\sigma$ be a
  domain size function. Then the set of models of $(\psi, \sigma \circ \rho)$ is
  equal to the set of models of $(\phi, \sigma)$.
\end{theorem}
\begin{proof}
  We first show that the right-hand side of \cref{eq:subset} is the same for
  models of both $(\phi, \sigma)$ and $(\psi, \sigma \circ \rho)$.
  \Cref{alg:trycache} ensures that $\psi \equiv \phi$ up to domains. In
  particular, this means that $\Preds(\psi) = \Preds(\phi)$. The square in
  \begin{equation}\label[diagram]{diagram:ref}
    \begin{tikzcd}
      \Preds(\psi) \ar[equal]{r} \ar[d, swap, "\pi_{\psi}"] & \Preds(\phi) \ar[d, "\pi_{\phi}"] \\
      {\Doms(\psi)}^{\ast} \ar[r, "\rho"] & {\Doms(\phi)}^{\ast} \ar[r, "\sigma"] & \mathbb{N}_{0}^{\ast}
    \end{tikzcd}
  \end{equation}
  commutes by \cref{def:formula} and the definition of $\rho$. In other words,
  for any predicate $\texttt{p}/n \in \Preds(\psi) = \Preds(\phi)$, function
  $\rho$ translates the domains associated with $\texttt{p}$ in $\psi$ to the
  domains associated with $\texttt{p}$ in $\phi$. Let
  $\pi_{\phi}(\texttt{p}) = {(d_{i})}_{i=1}^{n}$ and
  $\pi_{\psi}(\texttt{p}) = {(e_{i})}_{i=1}^{n}$ for some domains
  $d_{i} \in \Doms(\phi)$ and $e_{i} \in \Doms(\psi)$. Since
  $\rho(e_{i}) = d_{i}$ for all $i = 1, \dots, n$ by the commutativity in
  \cref{diagram:ref}, we get that
  \begin{equation}\label{eq:equality}
    \prod_{i=1}^{n}[\sigma(d_{i})] = \prod_{i=1}^{n}[\sigma \circ \rho(e_{i})]
  \end{equation}
  as required. Certainly, any subset of the left-hand side of \cref{eq:equality}
  (e.g., $\mathfrak{M}(\texttt{p})$, where $\mathfrak{M}$ is a model of
  $(\phi, \sigma)$) is a subset of the right-hand side and vice versa. Since
  $\phi$ and $\psi$ semantically only differ in what domains they refer to,
  every model of $\phi$ satisfies $\psi$ and vice versa, completing the proof.
\end{proof}

@@

In this appendix, we list the exact function definitions produced by
\textsc{Crane} for all of the problem instances in Section~6 both before and
after algebraic simplification (excluding multiplications by one). The
correctness of all of them has been checked by identifying suitable base cases
and verifying the numerical answers across a range of domain sizes.

\begin{enumerate}
  \item $\Theta(m)$ solution for counting $\Gamma \to \Gamma$ functions:
  \[
    f(m) = {\left(-1 + \sum_{l=0}^{m} \binom{m}{l} [l < 2]\right)}^{m} = m^{m}.
  \]
  \item $\Theta(m^3 + n^3)$ solution for counting $\Gamma \to \Delta$
  surjections:
  \begin{align*}
    f(m, n) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k}\\
               &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
    ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k} {(k+1)}^{l}.
  \end{align*}
  \item $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ surjections:
  \begin{align*}
    f(m) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k}\\
            &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
    ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k} {(k+1)}^{l}.
  \end{align*}
  \item $\Theta(mn)$ solution for counting $\Gamma \to \Delta$ injections and
  partial injections (with different base cases):
  \begin{align*}
    f(m, n) ={}& \sum_{l=0}^m \binom{m}{l} [l<2] f(m-l, n-1)\\
    ={}& f(m, n-1) + mf(m-1, n-1).
  \end{align*}
  \item $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ injections:
  \begin{align*}
    f(m) ={}& \sum_{l=0}^{m} \binom{m}{l} {(-1)}^{m-l} g(m, l); \\
    g(m, l) ={}& \sum_{k=0}^{l} \binom{l}{k} [k < 2] g(m - 1, l - k)\\
    ={}& g(m - 1, l) + lg(m - 1, l - 1).
  \end{align*}
  % \item $\Theta(m^{2})$ solution for counting $\Gamma \to \Gamma$ partial
  %       injections:
  % \begin{align*}
  %   f(m) ={}& f(m - 1) + \sum_{l = 0}^{m} \binom{m}{l} \sum_{k = 0}^{m} \binom{m}{k}\\
  %           &[1 \le l < 3] [k < 2] 2^{m - l} g(m - l, m - k - 1)\\
  %   ={}& f(m - 1) + m2^{m-1}g(m-1, m-1)\\
  %   +{}& m^{2}2^{m-1}g(m-1, m-2)\\
  %   +{}& \binom{m}{2}2^{m-2}g(m-2, m-1) \\
  %   +{}& \binom{m}{2}m2^{m-2}g(m-2, m-2)\\
  %   g(i, j) ={}& \sum_{h = 0}^{j} \binom{j}{h} [h < 2] g(i - 1, j - h)\\
  %   ={}& g(i-1, j) + jg(i-1, j-1).
  %   \end{align*}
  \item $\Theta(m)$ solution for counting $\Gamma \to \Delta$ bijections:
  \[
    f(m, n) = mf(m-1, n-1).
  \]
\end{enumerate}

>>>>

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usepackage{siunitx}
\usepackage{forest}
\usepackage[inline]{enumitem}
\usepackage{xr}
\externaldocument{paper}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}[theorem]
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand{\predicate}{\texttt{\textup{p}}}

@@

We begin with a few definitions that formalise the notion of a model. For any
set $S$, let $2^{S}$ denote its power set and
$S^{\ast} \coloneqq \bigcup_{i=0}^{\infty} S^{i}$ the set of tuples---of any
finite length---of elements of $S$. For example,
${\{\,a, b\,\}}^{\ast} = \{\, (), (a), (b), (a, a), \dots \,\}$. For any
$n \in \mathbb{N}_{0}$, let $[n] \coloneqq \{\, 1, \dots, n \,\}$, e.g.,
$[0] = \emptyset$, and $[2] = \{\, 1, 2 \,\}$. Extending the notation for
substitution introduced in \cref{sec:recprelims}, for a formula $\phi$, constant
$c$, and set of variables $V$, let $\phi[c/V]$ denote $\phi$ with all
occurrences of all variables in $V$ replaced with $c$. Let $\Preds$ be the
function that maps any clause or formula to the set of predicates used within.
With each formula $\phi$ we associate a map
$\pi_{\phi}\colon \Preds(\phi) \to {\Doms(\phi)}^{\ast}$ s.t., for each
predicate $\predicate/n \in \Preds(\phi)$, we have that
$\pi_{\phi}(\predicate) \in {\Doms(\phi)}^{n}$. This map makes explicit the idea
that each $n$-ary predicate is associated with an $n$-tuple of domains. In the
case of formula $\phi$ from \cref{example:first},
$\pi_{\phi} = \{\, \predicate \mapsto (\Gamma, \Delta) \,\}$.

\begin{definition}\label{def:model}
  Let $\phi$ be a formula and $\sigma$ a domain size function, A \emph{model} of
  $(\phi, \sigma)$ is a map
  $\mathfrak{M}\colon \Preds(\phi) \to 2^{\mathbb{N}_{0}^{\ast}}$ s.t.\ the
  following two conditions are satisfied.
  \begin{enumerate}
    \item For all predicates $\predicate/n \in \Preds(\phi)$ with
          $\pi_{\phi}(\predicate) \coloneqq {(\Delta_{i})}_{i=1}^{n}$ for some
          domains $\Delta_{i} \in \Doms(\phi)$,
    \begin{equation}\label{eq:subset}
      \mathfrak{M}(\predicate) \subseteq \prod_{i=1}^{n}[\sigma(\Delta_{i})].\footnote{For simplicity, \cref{def:model} ignores constants. To include constants, one would replace $[\sigma(\Delta_{i})]$ with a set that contains all constants associated with domain $\Delta_{i}$, extended with enough new elements to make its cardinality $\sigma(\Delta_{i})$.}
    \end{equation}
    \item $\mathfrak{M} \models \phi$ according to the (non-exhaustive)
          recursive definition below. Let $\chi$ and $\psi$ be formulas,
          $\Delta$ a domain, $c \in \Delta$ a constant, $\predicate/n$ a
          predicate associated with domains $\Delta_{1},\dots,\Delta_{n}$, and
          $t \in \prod_{i=1}^{n} \Delta_{i}$ a tuple of constants.
    \begin{itemize}
      \item $\mathfrak{M} \models \predicate(t)$ if and only if (iff)
            $t \in \mathfrak{M}(\predicate)$.
      \item $\mathfrak{M} \models \neg\chi$ iff $\mathfrak{M} \not\models \chi$.
      \item $\mathfrak{M} \models \chi \land \psi$ iff
            $\mathfrak{M} \models \chi$ and $\mathfrak{M} \models \psi$.
      \item $\mathfrak{M} \models \chi \lor \psi$ iff
            $\mathfrak{M} \models \chi$ or $\mathfrak{M} \models \psi$.
      \item $\mathfrak{M} \models (\forall X \in \Delta\text{. }\chi)$ iff
            $\mathfrak{M} \models \chi[x/\{\, X \,\}]$ for all constants
            $x \in \Delta$.
      \item $\mathfrak{M} \models (\forall X \in \Delta\text{.
            } X \ne c \Rightarrow \chi)$ iff
            $\mathfrak{M} \models \chi[x/\{\, X \,\}]$ for all
            $x \in \Delta \setminus \{\, c \,\}$.
      \item $\mathfrak{M} \models (\forall X,Y \in \Delta\text{.
            } X \ne Y \Rightarrow \chi)$ iff
            $\mathfrak{M} \models \chi[x/\{\, X \,\}][y/\{\, Y \,\}]$ for all
            $x,y \in \Delta$ s.t. $x \ne y$.
    \end{itemize}
  \end{enumerate}
\end{definition}

@@

\begin{theorem}[Correctness of $\Reff$]
  Let $\phi$ be the formula used as input to \cref{alg:trycache}. Let $\psi$ be
  any formula selected on \cref{line:selectformula} of the algorithm s.t.\
  $\rho \ne {\normalfont\texttt{null}}$ on \cref{line:rho}. Let $\sigma$ be a
  domain size function. Then the set of models of $(\psi, \sigma \circ \rho)$ is
  equal to the set of models of $(\phi, \sigma)$.
\end{theorem}
\begin{proof}
  We first show that the right-hand side of \cref{eq:subset} is the same for
  models of both $(\phi, \sigma)$ and $(\psi, \sigma \circ \rho)$.
  \Cref{alg:trycache} ensures that $\psi \equiv \phi$ up to domains. In
  particular, this means that $\Preds(\psi) = \Preds(\phi)$. The square in
  \begin{equation}\label[diagram]{diagram:ref}
    \begin{tikzcd}
      \Preds(\psi) \ar[equal]{r} \ar[d, swap, "\pi_{\psi}"] & \Preds(\phi) \ar[d, "\pi_{\phi}"] \\
      {\Doms(\psi)}^{\ast} \ar[r, "\rho"] & {\Doms(\phi)}^{\ast} \ar[r, "\sigma"] & \mathbb{N}_{0}^{\ast}
    \end{tikzcd}
  \end{equation}
  commutes by \cref{def:formula} and the definition of $\rho$. In other words,
  for any predicate $\predicate/n \in \Preds(\psi) = \Preds(\phi)$, function
  $\rho$ translates the domains associated with $\predicate$ in $\psi$ to the
  domains associated with $\predicate$ in $\phi$. Let
  $\pi_{\phi}(\predicate) = {(\Delta_{i})}_{i=1}^{n}$ and
  $\pi_{\psi}(\predicate) = {(\Gamma_{i})}_{i=1}^{n}$ for some domains
  $\Delta_{i} \in \Doms(\phi)$ and $\Gamma_{i} \in \Doms(\psi)$. Since
  $\rho(\Gamma_{i}) = \Delta_{i}$ for all $i = 1, \dots, n$ by the commutativity
  in \cref{diagram:ref}, we get that
  \begin{equation}\label{eq:equality}
    \prod_{i=1}^{n}[\sigma(\Delta_{i})] = \prod_{i=1}^{n}[\sigma \circ \rho(\Gamma_{i})]
  \end{equation}
  as required. Certainly, any subset of the left-hand side of \cref{eq:equality}
  (e.g., $\mathfrak{M}(\predicate)$, where $\mathfrak{M}$ is a model of
  $(\phi, \sigma)$) is a subset of the right-hand side and vice versa. Since
  $\phi$ and $\psi$ semantically only differ in what domains they refer to,
  every model of $\phi$ satisfies $\psi$ and vice versa, completing the proof.
\end{proof}

@@

Here we list the exact function definitions produced by \textsc{Crane} for all
of the problem instances in Section~6 both before and after algebraic
simplification (excluding multiplications by one). The correctness of all of
them has been checked by identifying suitable base cases and verifying the
numerical answers across a range of domain sizes.

\begin{enumerate}
  \item A $\Theta(m)$ solution for counting $\Gamma \to \Gamma$ functions:
        \[
        f(m) = {\left(-1 + \sum_{l=0}^{m} \binom{m}{l} [l < 2]\right)}^{m} = m^{m}.
        \]
  \item A $\Theta(m^3 + n^3)$ solution for counting $\Gamma \to \Delta$
        surjections:
        \begin{align*}
          f(m, n) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k}\\
                     &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
          ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l}\\
                     &\sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k} {(k+1)}^{l}.
        \end{align*}
  \item A $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ surjections:
        \begin{align*}
          f(m) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k}\\
                  &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
          ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l}\\
                  &\sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k} {(k+1)}^{l}.
        \end{align*}
  \item A $\Theta(mn)$ solution for counting $\Gamma \to \Delta$ injections and
        partial injections (with different base cases):
        \begin{align*}
          f(m, n) ={}& \sum_{l=0}^m \binom{m}{l} [l<2] f(m-l, n-1)\\
          ={}& f(m, n-1) + mf(m-1, n-1).
        \end{align*}
  \item A $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ injections:
        \begin{align*}
          f(m) ={}& \sum_{l=0}^{m} \binom{m}{l} {(-1)}^{m-l} g(m, l); \\
          g(m, l) ={}& \sum_{k=0}^{l} \binom{l}{k} [k < 2] g(m - 1, l - k)\\
          ={}& g(m - 1, l) + lg(m - 1, l - 1).
        \end{align*}
  \item A $\Theta(m)$ solution for counting $\Gamma \to \Delta$ bijections:
        \[
        f(m, n) = mf(m-1, n-1).
        \]
  \item A $\Theta(l+mn)$ solution for counting $|\Lambda|$ partial injections
        $\Gamma \to \Delta$:
        \begin{align*}
          f(l, m, n) ={}& {g(m, n)}^{l}\\
          g(m, n) ={}& \sum_{k=0}^{n}\binom{n}{k} [k < 2] g(m-1, n-k)\\
          ={}& g(m-1, n) + ng(m-1, n-1).
        \end{align*}
\end{enumerate}
